---
Title: "Replication of Chou 2016 (Study 5) - Economic Insecurity Increases Physical Pain"
Author: "Sean Zion"
Date: "December 4th, 2017"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r Load Packages, echo=FALSE}
library(ggplot2)
library(tidyverse)
library(knitr)
library(ggthemes)
library(psych)
```

##Introduction

Prevalence of economic insecurity and physical pain have both increased dramatically in recent years. A series of studies by Chou, Parmar, and Galinsky (2016) report a causal link between the two, providing further evidence for a relationship between broad social trends and physical health. This finding has important public health implications: pain has an economic cost of more than $635 billion and increased over 1000% in the past two decades. Connecting these two major social issues may critically shape public policy and should be investigated further. 

Chou and colleagues expand upon their finding by showing the causal link between economic insecurity and physical pain is mediated by feelings of lack of control. The psychological construct of lack of control has been implicated in numerous psychological and physiological states. The goal of this replication is to replicate the results of study 5, which reports a causal link between their proposed mediating variable, lack of control, and their dependent variable, self-reported physical pain. While this does not replicate the main finding of the paper, the relationship between lack of control and physical pain is an important component of Chou's proposed model.

##Methods

###Power Analysis

100 participants were recruited for Study 5. A simple t-test was used in their analysis of the data (t(98=2.04; p=0.04)), which yeilded an effect size of d=0.41. A post-hoc power analysis revealed that the study was powered at 53%. To acheive 80% power, an a priori power analysis conduced using G Power suggested a sample of 156 subjects across two groups.

###Planned Sample

156 adult subjects will be recuited via Amazon's Mechanical Turk and will be randomly assigned to either the high control or low control condition. mTurk workers will be recruited from the United States with a hit rate >90%. Any participants who does not provide consent (responds "disagree" to the question 'consent') will be excluded from the survey. All other participants will be included. 

###Materials

Participants were presented with the same questions as in Study 5 of Chou et al., 2016. 


###Procedure

Participants began by reading the IRB header and were asked to agree / disagree to contiue. They were then randomized, in a 1:1 ratio, to either a high control or low control condition. In the high control condition, participants were asked to write about a time in their life when they felt in total control of a situation. They are allowed to choose the amount of time spent on this response. In the low-control condition, participants are asked to similarly describe a time in their life when they had no control over a situation. They are required to spend 1.5 minutes on this response before the page automatically advances. The purpose of this automatic advancement after 1.5 minutes is to reinforce the effect of the lack of control manipulation.

All participants are then asked to record the amount of physical pain they are currently experiencing, on a 0-100 slide scale. 

Subjects then completed a short, five-item lack of control measure, that was an additional scale added to the author's original procedure. See **Differences from Original Study** section for more details. 

Finally, participants responeded to a short demographics questionnaire in which they were asked their age, gender, employment status, and use of painkillers.

**Excerpt from paper:**

"Participants were asked to write about an autobiographical experience (Whitson & Galinsky, 2008). Half recalled a time when they lacked control. The other half recalled a time when they had complete control. 

After the recall task, participants were asked to “choose the overall pain level that best describes how much physical pain you are experiencing RIGHT NOW.” They responded using a visual slider scale (Portenoy & Kanner, 1996), on which 0 indicated no pain and 100 indicated the worst pain ever experienced.

Participants also reported their age, gender, and current employment status, and whether they were using painkillers."


**Replication Link**

https://stanforduniversity.qualtrics.com/jfe/form/SV_cOKVcCFKgzvOxxj

 
###Analysis Plan

**Analysis Steps**

1) Data Cleaning: Summarize the data so that each participant has a separate column for the manipulation condition (low control / high control), response to pain questionnaire, and control variables / demographics. The mean score for the five-item lack of control questionnaire will be calculated for each participant as the average of the five individual responses.

2) Analysis: 
- Calculate descriptive statistics for the two groups. Mean and SD will be calculated for pain rating, five-item lack-of-control questionnaire, and control variables / demographics.

- Conduct a t-test for mean difference in pain ratings between the two manipulation groups. 

-Conduct a regression model in which in pain ratings between the two manipualtion groups controlling for age, gender, and painkiller use. 


###Differences from Original Study

**Sample differences**

None: the original paper used a sample of 100 mTurk workers in 2015. It can be assumed that a similar sample would be obtained with the replication.

**Setting differences**

Slight: We will be paying participants 60 cents for their time. The original study paid 41 cents. Given the minor adjustment in cost, we do not beleive this will have any effect on the results.

**Procedure differences**

The replication was conducted exactly as outlined in Study 5 of Chou et al., with the addition of a short, five-item scale that captures respondents’ psychological sense of lacking control (Lachman & Weaver, 1998). This scale was used in study 4 of the paper, and serves as an aditional measure of control to be used in additional analyses of the overall hypothesis presented by Chou and colleagues.

Two additional control variable questions were added to the end of the survey. Following the question regarding use of pain killers, participants who responded yes to this were routed to two additional questions about painkiller type and use frequency. These were added so that additonal analyses could be conducted on the subpopulation of participants who were taking painkilling drugs to see if this a self-identified clinical sample would experience the same pattern of results. 


**Analysis plan differences**

As identified above, we are collecting an additional measure of control - the Lachman & Weaver five-item control scale. First, I hope to use this scale as a manipulation check to see if the manipualtion result in a significant mean difference in this scale. The goal of the high/low control manipulation is to induce a sense of control or a sense of lack of control. The authors do not indicate any check to ensure this manipulation was effective in the original paper. The manipulation may - rather than altering sense of control - cause participants to reflect on experiences with different affective valences. Instances of low control may simply be more upsetting, and it is this affective component that drives difference in pain ratings. 


##Methods Addendum (Post Data Collection)

- You can comment this section out prior to final report with data collection.

#### Actual Sample
- Sample size, demographics, data exclusions based on rules spelled out in analysis plan

#### Differences from pre-data collection methods plan
- Any differences from what was described as the original plan, or “none”.


##Results

###Data preparation

**Load the data**
```{r Import Data File, echo = FALSE}
d <- read.csv("~/Chou2016/Final/chou2016_final.csv")
```


```{r Clean & Filter Data, echo=FALSE}
d2 <- d[-c(1:2), ] %>%
  select(consent:painkiller_freq)
```

> First let's remove rows one and two since they do not contain data. Then we can select only the columns of interest.

```{r Tidy Data, echo=FALSE}
d2['manipulation'] <- paste0(d2$high_control, d2$low_control)
d2['condition'] <- ifelse(as.character(d2$high_control)==as.character(d2$manipulation), 'high_control', 'low_control')
d2 = d2[ , -which(names(d2) %in% c("high_control","low_control"))]
d2 = d2[, c(ncol(d2),ncol(d2)-1,seq(1,ncol(d2)-2))]
d2$pain_rate_1 <- as.numeric(d2$pain_rate_1)
```

> To view the data in a format that is easier to understand, we have first organized in 'Tidy Data' format. Next, a column was created to indicate the manipulation condition to which the participant was randomized (high_control / low_control).

**Descriptive Statistics - Pain Ratings**

*Across All Subjects* 
```{r Descriptive Statistics 1 - Pain Rating, echo=FALSE}
print(mean(d2$pain_rate_1))
describe(d2$pain_rate_1)
```

> The mean rating of current pain across all 158 subjects was 18.96 (SD=16.82) on a 100 point slide scale.

*Subjects in High Control Condition* 
```{r Descriptive Statistics 2 - Pain Rating, echo=FALSE}
#Pain Ratings for Subjects in High Control Condition:
print(mean(subset(d2$pain_rate_1,d2$condition=='high_control')))
describe(subset(d2$pain_rate_1,d2$condition=='high_control'))
```

> The mean pain rating of current pain for the 86 subjects in the high control condition was 19.65 (SD=15.75).

*Subjects in Low Control Condition*
```{r}
# Pain Ratings for Subjects in Low Control Condition:
print(mean(subset(d2$pain_rate_1,d2$condition=='low_control')))
describe(subset(d2$pain_rate_1,d2$condition=='low_control'))
```

> The mean pain rating of current pain for the 72 subjects in the low control condition was 18.14 (SD=18.09).

**Descriptive Statistics - Five Item Control Scale Ratings**

*Across All Subjects* 
```{r Descriptive Statistics 1 - Control}
##Five Item Control
controls = c('control_1', 'control_2', 'control_3', 'control_4', 'control_5')
for (control in controls){
  d2[, control] = as.numeric(as.character(d2[, control]))
}
d2['five_item_control'] = rowMeans(d2[, controls])

#Across All Subjects
print(mean(d2$five_item_control))
describe(d2$five_item_control)
```

> The mean score on the five-item lack of control scale for all 158 subjects was 3.71 (SD=0.89).

*Subjects in High Control Condition* 
```{r Descriptive Statistics 2 - Control, echo=FALSE}
#Control Score for Subjects in High Control Condition:
print(mean(subset(d2$five_item_control, d2$condition=='high_control')))
describe(subset(d2$five_item_control, d2$condition=='high_control'))
```

> The mean score on the five-item lack of control scale for all the 86 subjects in the high control condition was 3.70 (SD=0.88).

*Subjects in Low Control Condition*
```{r Descriptive Statistics 3 - Control, echo=FALSE}
#Control Score for Subjects in Low Control Condition:
print(mean(subset(d2$five_item_control, d2$condition=='low_control')))
describe(subset(d2$five_item_control, d2$condition=='low_control'))
```

> The mean score on the five-item lack of control scale for all the 86 subjects in the high control condition was 3.73 (SD=0.90).

###Confirmatory analysis

*Primary Analysis*
```{r T-Test}
#Main Analysis - t test 
t.test(subset(d2$pain_rate_1, d2$condition=='high_control')
       , subset(d2$pain_rate_1, d2$condition=='low_control')
       , alternative = "two.sided", var.equal = FALSE)
```

> A Welch Two Sample t-test of the mean pain rating between the two manipulation conditions returned a t=0.55; df=141.97; p=0.58; 95% CI of the mean difference between conditions was -3.87 - 6.90.

*Linear Model controlling for age and gender*

> To determine the influence of potentially mediating variables, a linear model has been constructed. In addition to the condition to which each subject has been assigned, this model adds age, gender, employment status, and painkiller use to the set of variables predicting pain rating.

```{r}
#Make sure class is correct for each variable
d2$age[d2$age==""] <- NA
d2$gender[d2$gender==""] <- NA
d2$employment[d2$employment==""] <- NA
d2$painkiller[d2$painkiller==""] <- NA
d2$condition = as.factor(d2$condition)
d2$age = as.numeric(d2$age)
d2$gender = as.factor(d2$gender)
d2$employment = as.factor(d2$employment)
d2$painkiller = as.factor(d2$painkiller)




lin_mod = lm(pain_rate_1 ~ condition + age + gender + employment + painkiller, data = d2)
summary(lin_mod)
```

> A linear model controlling for age, gender, employment status, and painkiller use was constructed. Painkiller use was the only factor that was significantly positively related to pain rating (b=10.76; SE=3.38; t-value=3.18; p<0.01). As with the t-test above, the condition to which the subject was assigned (high control / low control) was not a significant variable in the model.


> Next, to visualize the data, we have constructed a boxplot of pain ratings across the two manipulation conditions (low control & high control).

```{r Mean Pain Rating Across Conditions - Boxplot}
ggplot(d2, aes(x = condition, y = pain_rate_1)) +
  geom_boxplot(fill = "#4271AE", colour = "#1F3552", alpha = 0.7, outlier.colour = "#1F3552", outlier.shape = 20) +
  scale_y_continuous(name = "Mean Pain Rating", breaks = seq(0, 75, 15), limits=c(0, 75)) +
  scale_x_discrete(name = "Control Condition") +
  ggtitle("Mean Pain Rating Across Control Conditions") +
  stat_summary(fun.y=mean, colour='red', geom='point') +
  theme_bw() + theme(plot.title = element_text(hjust = 0.5))
```

> A boxplot of pain ratings across the two control conditions is presented. Pain ratings were assessed on a 0-100 scale, however, for purposes of visualization, a 0-75 point scale is shown. The mean for each group is marked in red for each group on the boxes.

###Exploratory analyses


```{r}
d3 <- data.frame(filter(d2, painkiller==2))

print(mean(subset(d3$pain_rate_1,d3$condition=='high_control')))
describe(subset(d3$pain_rate_1,d3$condition=='high_control'))

print(mean(subset(d3$pain_rate_1,d3$condition=='low_control')))
describe(subset(d3$pain_rate_1,d3$condition=='low_control'))

t.test(subset(d3$pain_rate_1, d3$condition=='high_control')
       , subset(d3$pain_rate_1, d3$condition=='low_control')
       , alternative = "two.sided", var.equal = FALSE)

```


```{r Pain Rating as a Function of 5-Item Control Scale}

#Pain Rating as a Function of 5-Item Control Scale
ggplot(d2, aes(x=five_item_control, y=pain_rate_1, color=condition))+
  geom_point() +
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE) +
  labs(title="Pain Rating as a Function of 5-Item Control Scale",
       x="Mean 5-Item Control", y = "Pain Rating") + 
  theme_bw() + theme(plot.title = element_text(hjust = 0.5))

```






##Discussion

###Summary of Replication Attempt

###Commentary

